---
title: "Practical Machine Learning"
output: html_document
---
#Introduction
The datasets used in this project are from 
http://groupware.les.inf.puc-rio.br/har website.
This data contains the measurement of excercise activity  of the participants.
Different measurements from Arm,Forearm,Belt and Dumbell sensors were used to 
record this. The goal of this project is to take the data given in the training
dataset , build a model to predict the classe variable and use it to finally 
predict classe variable in the testing dataset. 
training dataset - 
 https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
testing dataset -
 https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
Outcome variable classe tells whether the participant correctly performed the 
excercise as specified (Classification A) or 
did somethings wrong (Classifications B,C,D,E)
```{r,echo=FALSE, message=FALSE,warning=FALSE}
library(caret)
library(doMC)
library(randomForest)
library(RANN)
```
#Data Processing 
Downloading the training and testing files and loading data
```{r}
download.file(
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
        destfile = "WtLiftingExcerciseTraining.csv", method = "curl")
download.file(
        "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
        destfile = "WtLiftingExcerciseTesting.csv", method = "curl")

actualTrainingData <- read.csv("WtLiftingExcerciseTraining.csv",header = TRUE,
                               sep=",",na.strings=c("NA","","#DIV/0!"),
                               stringsAsFactors = FALSE)

testing <-read.csv("WtLiftingExcerciseTesting.csv",,header = TRUE,sep=",",
                   na.strings=c("NA","","#DIV/0!"),stringsAsFactors = FALSE)
```
```{r}
dim(actualTrainingData)
dim(testing)
str(actualTrainingData)
```
Selecting interested columns and ignoring the statiscal computed variables 
such as avg , var , stddev etc
```{r}
trainingInterestedCols <- actualTrainingData[, -grep('(X|raw_timestamp_part_1|raw_timestamp_part_2|cvtd_timestamp|^avg|^var|^stddev|^min|^max|^amplitude|^kurtosis|^skewness)', colnames(actualTrainingData))]
```
Converting all interested Character variables to factor variables before training the model
```{r}
trainingInterestedCols$user_name <- as.factor(trainingInterestedCols$user_name)
testing$user_name <- as.factor(testing$user_name)
trainingInterestedCols$classe <- as.factor(trainingInterestedCols$classe)
```
using Caret Package , Splitting  the training data into  training and 
validation datasets.  A 70/30 split is used for this.
Validation dataset is set aside to check the model accuracy and estimate
out of sample error etc.
```{r}
inTrain <- createDataPartition(y=trainingInterestedCols$classe,p=0.7,list = FALSE)
training <- trainingInterestedCols[inTrain,]
validation <- trainingInterestedCols[-inTrain,]
```
removing columns that has all NA values 
```{r}
trainingNonNA <- training[ , !apply( training , 2 , function(x) all(is.na(x)) )]
```
removing non zero var variables (i.e here new_window)
```{r}
training <- trainingNonNA[, -nearZeroVar(trainingNonNA)]
```
#Modeling
Trained the model using the Caret package
and also utilized the parallel processing capabilities of doMC package
To find the optimal model and for cross validatiion,used cv method with 5 folds.
```{r}
registerDoMC(cores = 24)
modelFit <- train(classe ~ ., 
                          data = training, 
                          method = "rf",
                          trControl = trainControl(method = "cv", 
                                                   number = 5)) 
modelFit
summary(modelFit)
plot(modelFit)
```
An optimal model with mtry = 30 and best accuracy 99.69% is chosen.

#Prediction and Analysis
Applying the model to validation dataset to check accuracy or fine tune model
if necessary.
```{r}
validationPred <- predict(modelFit,validation)
cm<-confusionMatrix(validationPred,validation$classe)
cm
```

`r modelAccuracy <- cm$overall['Accuracy']`
Based on this we can see that Accuracy is `r modelAccuracy`
(>99%) and
we expect the out of sample error to be 1 - Accuracy = `r 1-modelAccuracy`


Finally applied the model to testing dataset to predict the classe variable.
```{r}
testingPred <- predict(modelFit,testing)
testingPred
```

Writing the final prediction results to output files
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(testingPred)
```


